---
title: "Data Simulation Project"
author: "Uma Ramasubramanian"
date: "May 7, 2017"
output: html_document
---


```{r}
set.seed(101)
generateDataset<-function(N){
  data.frame(matrix(0, nrow = N, ncol = 11))
  age<-0
  canc<-0
  cvd<-0
  diab<-0
  hbp<-0
  education<-0
  smoke<-0
  functional.limitation<-0
  physical.activity<-0
  bmi<-0
  sitting<-0
  data.frame(age,canc,cvd,diab,hbp,education,smoke,functional.limitation,physical.activity,bmi,sitting)
  age<-rnorm(N,55.6,5.4)
  education<-factor(sample(0:5, N, replace=TRUE, prob=c(0.07, 0.128, 0.104, 0.172,0.21,0.317)),ordered=T, 
                    levels=c('0','1','2','3','4','5'))
  smoke<-rbinom(N,1,0.48)
  
  functional.limitation<-factor(sample(0:3, N, replace=TRUE, prob=c(0.445,0.205, 0.169, 0.181)),ordered=T,
                                levels= c('0','1','2','3'))
  no<-which(functional.limitation==0)
  length.no<-length(no)
  mild<-which(functional.limitation==1)
  length.mild<-length(mild)
  mod<-which(functional.limitation==2)
  length.mod<-length(mod)
  severe<-which(functional.limitation==3)
  length.severe<-length(severe)
  
  sitting <- factor( rep("level4",N), ordered=T,levels =c("level1", "level2", "level3", "level4"))
  mod1<-sample(mod,round(length(mod)*0.35),replace = FALSE)
  no1<-sample(no,round(length.no*0.30),replace = FALSE)
  mild1<-sample(mild,round(length.mild*0.20),replace = FALSE)
  sitting[c(no1,mild1,mod1)]<-"level1"

  
  no2<-sample(subset(no, !(no %in% no1)),round(length.no*0.30),replace = FALSE)
  mild2<-sample(subset(mild, !(mild %in% mild1)),round(length.mild*0.25),replace = FALSE)
  mod2<-sample(subset(mod,!(mod %in% mod1)),round(length.mod*0.10),replace = FALSE)
  sitting[c(no2,mild2,mod2)]<-"level2"
  
  mod3<-sample(subset(mod, !(mod %in% c(mod1,mod2)),round(length.mod*0.75),replace = FALSE))
  no3<-sample(subset(no, !(no %in% c(no1,no2))),round(length.no*0.35),replace = FALSE)
  sitting[c(mod3,no3)]<- "level3"

  logistic <- function(t) 1 / (1 + exp(-t))
  canc <- runif(length(age))< .05*logistic((age-50)/10) + .065*logistic((as.numeric(sitting)-3)/2)
  cvd<- runif(length(age))< .06*logistic((age-45)/2) + .07*logistic((as.numeric(sitting)-3)/2)
  diab <- runif(length(age))< .07*logistic((age-50)/10) + .086*logistic((as.numeric(sitting)-3)/2)
  hbp <- runif(length(age))< .4*logistic((age-45)/10) +.06*logistic((as.numeric(sitting)-3)/2)

  
  physical.activity <- factor( rep("veryhigh",N), ordered=T,levels =c("sed", "low", "suff", "high", "veryhigh"))
  low.mod1<-sample(mod,round(length(mod)*0.8),replace = FALSE)
  low.no1<-sample(no,round(length.no*0.03),replace = FALSE)
  low.mild1<-sample(mild,round(length.mild*0.05),replace = FALSE)
  physical.activity[c(low.no1,low.mild1,low.mod1)]<-"low"
  
  suff.no2<-sample(subset(no, !(no %in% low.no1)),round(length.no*0.28),replace = FALSE)
  suff.mild2<-sample(subset(mild, !(mild %in% low.mild1)),round(length.mild*0.45),replace = FALSE)
  suff.mod2<-sample(subset(mod,!(mod %in% low.mod1)),round(length.mod*0.05),replace = FALSE)
  physical.activity[c(suff.no2,suff.mild2,suff.mod2)]<-"suff"
  
  high.no3<-sample(subset(no, !(no %in% c(no1,no2)),round(length.no*0.1),replace = FALSE))
  high.mild3<-sample(subset(mild, !(mild %in% c(low.mild1,suff.mild2)),round(length.mild*0.1),replace = FALSE))
  physical.activity[c(high.no3,high.mild3)]<-"high"

  severe.sed<-sample(severe,length.severe,replace = FALSE)
  severe.no4<-sample(subset(no, !(no %in% c(low.no1,suff.no2,high.no3)),round(length.no*0.1),replace = FALSE))
  physical.activity[c(severe.sed,severe.no4)]<-"sed"
  
  
  bmi<- factor( rep("obese",N), ordered=T,levels =c("underweight","normalweight", "overweight","obese"))
  
  sed<-which(physical.activity=="sed")
  length.sed<-length(sed)
  low<-which(physical.activity=="low")
  length.low<-length(low)
  suff<-which(physical.activity=="suff")
  length.suff<-length(suff)
  high<-which(physical.activity=="high")
  length.high<-length(high)
  veryhigh<-which(physical.activity=="veryhigh")
  length.veryhigh<-length(veryhigh)

  high1<-sample(high,round(length.high*0.01),replace = FALSE)
  veryhigh1<-sample(veryhigh,round(length.veryhigh*0.015),replace = FALSE)
  bmi[c(high1,veryhigh1)]<-"underweight"
  
  high2<-sample(subset(high, !(high %in% high1)),round(length.high*0.75),replace = FALSE)
  veryhigh2<-sample(subset(veryhigh, !(veryhigh %in% veryhigh1)),round(length.veryhigh*0.5),replace = FALSE)
  low2<-sample(low,round(length.low*0.05),replace = FALSE)
  sed2<-sample(sed,round(length.sed*0.05),replace = FALSE)
  suff2<-sample(suff,round(length.suff*0.25),replace = FALSE)
  bmi[c(high2,veryhigh2,low2,sed2,suff2)]<-"normalweight"
  
  
  sed3<-sample(subset(sed, !(sed %in% sed2),round(length.sed*0.1),replace = FALSE))
  suff3<-sample(subset(suff, !(suff %in% c(suff2)),round(length.suff*0.10),replace = FALSE))
  bmi[c(sed3,suff3)]<-"overweight"
  
  data.frame(age,canc,cvd,diab,hbp,education,smoke,functional.limitation,physical.activity,bmi,sitting)
}
```
```{r}
data<-generateDataset(50000)
```
```{r}
# Add new column counting the number of chronic diseases reported
data$chronic.disease<-rowSums(data=='TRUE')
```
```{r}
summary(data)
```


The data simulated looks close to the researchers data. Some additional noise was added to encounter for patient reporting the values of functional limitation, sitting time and physical activity level in error.


```{r}
library(ggplot2)
p<-ggplot(data, aes(x = as.numeric(physical.activity))) + geom_histogram(aes(fill=bmi),position = 'dodge', bins = 10)
p+labs(title="Distribution of BMI with physical activity",x="Physical Activity")
```

Most of population, who is low in  physical activity tend to be overweight or obese. There is distibution of overweight in other levels as well to encounter for error in reporting of the level og physical activity, weight and other factors like genetic make up, muscle mass contributing to higher BMI


```{r}
p<-ggplot(data, aes(x = as.numeric(functional.limitation))) + geom_histogram(aes(fill=physical.activity),position = 'dodge', bins = 10)
p+labs(title="Distribution of physical activity with functional limitation ",x="Functional limitation")

```

Higher level of functional limitation is assoicated with more sedentary lifestyle.However, there is a large proportion of the population that has no functional limitation and is sedentary.

```{r}
ggplot(data = data, aes(x = functional.limitation, y = physical.activity)) +
  geom_tile(aes(fill = sitting)) 

```

There is a high percentage of sedentary individuals in all levels of functional limitations. Most of those individuals have higher level of sitting as well.

```{r}
g<-ggplot(data,aes(chronic.disease)) + stat_count(aes(y = ((..count..)/sum(..count..))))
p<-g+ facet_wrap(~sitting)
p + labs (title= 'Distribution of Chronic Disease with level of sitting',y= 'Percentage')
```

As the level of sitting increases, the probability of the person having chronic diseases increases.
People with higher level of sitting have higher probability of getting 2 or more conditions

```{r}
data$sitting<-factor(data$sitting, ordered=FALSE)
model1<-glm(formula = canc~sitting, family = binomial(),data = data)
summary(model1)
```
The model on the whole is significant and level 4 of sitting is slightly significant to increase the odds of developing cancer by a factor of 0.3 and level 3 by a factor of 0.2

```{r}
model2<-glm(formula = diab~sitting + age, family = binomial(),data = data)
summary(model2)
```

Diabetes is highly correlated with all levels of sitting. On observing the coefficents we can see that the odds of developing diabetes increases as the level of sitting increases. The coefficents of higher level of sitting is more than age

```{r}
model3<-glm(formula = cvd~sitting , family = binomial(),data = data)
summary(model3)
```
Higher level of sitting is associated with higher probability of developing cardiovascular disease. The coefficent of level 4 sitting is almost 3 times as compared to level 2


```{r}
model4<-glm(formula = hbp~sitting, family = binomial(),data = data)
summary(model4)
```
Sitting for greater than 8 hours increases the odds of developing high blood pressure by a factor of 0.06

```{r}
data$chronic.disease<-as.factor(ifelse(data$chronic.disease==0,0,1))
model5<- glm(chronic.disease~sitting + age + physical.activity + functional.limitation,family = binomial(),data=data)
summary(model5)
```

Based on the above summary, we can see that the higher level of sitting and age are the significant variables increasing the odds of developing chronic disease

```{r}
model6<- glm(chronic.disease~sitting + age ,family = binomial(),data=data)
summary(model6)
```


The above is a better model with removal of the completely non relevant varaibles

Conclusion:
Higher level of sitting hours increases the odds of developing chronic conditions. The number of chronic conditions reported also increases as the amount of sitting hours increases. It is important for individuals with higher level of sitting to increase the level of physical activity, to decrease the chances of developing chronic conditions. Regular breaks every hour of sitting is found the decrease the risk significantly.
